Type cones of $\b{g}$-vector fans

### 2019-02-12 thirdPaper/draft/typeConeAssociahedra.tex
typos
Replaces \aps by \eps
improvements Section 2.2

In this section, we study the type cones of complete simplicial fans arising as normal fans of three families of generalizations of the associahedron: the generalized associahedra of finite type cluster algebras~\cite{FominZelevinsky-ClusterAlgebrasI, FominZelevinsky-ClusterAlgebrasII, FominZelevinsky-ClusterAlgebrasIV, HohlwegLangeThomas, HohlwegPilaudStella}, the gentle associahedra~\cite{PaluPilaudPlamondon-nonkissing}, and the graph associahedra~\cite{CarrDevadoss, Postnikov, FeichtnerSturmfels, Zelevinsky}. 
All these families contain the classical associahedra~$\Asso[n]$ constructed in~\cite{ShniderSternberg, Loday}.
We first describe these associahedra and their type cones as they are the prototypes of our constructions.

@@

@@

From these linear dependencies, we obtain a redundant description of the type cone of the fan~$\gvectorFan[n]$.

@@

For a seed~$\seed = (\B, \coefficients, \cluster)$ and a cluster variable~$x \in \seed$, the \defn{mutation} in direction~$x$ creates a new seed~$\mu_x(\seed) = \seed' = (\B', \coefficients', \cluster')$ where:
\begin{itemize}
\item the new cluster $\cluster'$ is obtained from $\cluster$ by replacing $x$ with the cluster variable $x'$ defined by the following \defn{exchange relation}:
\[
x x' = \positiveExponents{p_x} \prod_{{y \in \cluster, \; b_{xy}  > 0}} y^{b_{xy}} + \negativeExponents{p_x} \prod_{{y \in \cluster, \; b_{xy}  <0}} y^{-b_{xy}}
\]
and leaving the remaining cluster variables unchanged so that $\cluster \ssm \{x\} = \cluster' \ssm \{x'\}$.

@@

\para{Finite type}
%
In this paper, we only consider \defn{finite type} cluster algebras, \ie those having only finitely many cluster variables.
It turns out that these finite type cluster algebras were classified by S.~Fomin and A.~Zelevinsky~\cite{FominZelevinsky-ClusterAlgebrasII} using the Cartan-killing classification for crystallographic root systems.
Define the \defn{Cartan companion} of an exchange matrix $\B$ as the matrix~$\A{\B}$ given by~$a_{xy} = 2$ if~$x = y$ and~$a_{xy} = -|b_{xy}|$ otherwise.

@@

\para{Flips in the $\b{g}$-vector fan}
%
We now discuss the linear dependences between $\b{g}$-vectors of adjacent clusters.
The following proposition was stated in~\cite[Lem.~19]{HohlwegPilaudStella}.

@@

Note that this statement is unfortunately less precise than \cref{prop:exchangeablePairsAsso}. Indeed, which of the two possible linear dependences is satisfied by the $\b{g}$-vectors of~$\cluster \cup \cluster'$ depends on the initial exchange matrix~$\B_\circ$.
%In particular, the geometry of the $\b{g}$-vector fan~$\gvectorFan$ changes as $\B_\circ$ varies within a given mutation class.
However, it already ensures that the type cone~$\typeCone(\gvectorFan[\B_\circ])$ of the $\b{g}$-vector fan contains the polyhedral cone of~$\b{h} \in \R^{\variables}$ such that
\(
\b{h}_x + \b{h}_{x'} > \sum_{{y \in \cluster \cap \cluster', \; b_{xy} < 0}} -b_{xy}
\)
and
\(
% \quad\text{and}\quad
\b{h}_x + \b{h}_{x'} > \sum_{{y \in \cluster \cap \cluster', \; b_{xy} > 0}} b_{xy} \, \b{h}_y
\)
for any seeds~$(\B, \coefficients, \cluster)$ and~$(\B', \coefficients', \cluster')$ with~${\cluster \ssm \{x\} = \cluster' \ssm \{x'\}}$.
This is precisely where the vector~$\b{h}$ was chosen to ensure a realization in \cref{thm:generalizedAsso}.

>>>>

In this section, we study the type cones of complete simplicial fans arising as normal fans of three families of generalizations of the associahedron: the generalized associahedra of finite type cluster algebras~\cite{FominZelevinsky-ClusterAlgebrasI, FominZelevinsky-ClusterAlgebrasII, FominZelevinsky-ClusterAlgebrasIV, HohlwegLangeThomas, HohlwegPilaudStella}, the gentle associahedra~\cite{PaluPilaudPlamondon-nonkissing}, and the graph associahedra~\cite{CarrDevadoss, Postnikov, FeichtnerSturmfels, Zelevinsky}. 
All these families contain the classical associahedra~$\Asso[n]$ constructed in~\cite{ShniderSternberg, Loday}.

@@

We first describe the associahedra of~\cite{ShniderSternberg, Loday} and their type cones as they are the prototypes of our constructions.

\subsubsection{Associahedra and their normal fans}

@@

\subsubsection{Type cones of associahedra}

From the linear dependencies of \cref{prop:exchangeablePairsAsso}, we obtain a redundant description of the type cone of the fan~$\gvectorFan[n]$.

@@

For a seed~$\seed = (\B, \coefficients, \cluster)$ and a cluster variable~$x \in \cluster$, the \defn{mutation} in direction~$x$ creates a new seed~$\mu_x(\seed) = \seed' = (\B', \coefficients', \cluster')$ where:
\begin{itemize}
\item the new cluster $\cluster'$ is obtained from $\cluster$ by replacing $x$ with the cluster variable $x'$ defined by the following \defn{exchange relation}:
\[
x x' = \positiveExponents{p_x} \prod_{{y \in \cluster, \; b_{xy}  > 0}} y^{b_{xy}} + \negativeExponents{p_x} \prod_{{y \in \cluster, \; b_{xy}  <0}} y^{-b_{xy}}
\]
and leaving the remaining cluster variables unchanged so that $\cluster \ssm \{x\} = \cluster' \ssm \{x'\}$.

@@

\para{Finite type}
%
In this paper, we only consider \defn{finite type} cluster algebras, \ie those with a finite number~$N$ of cluster variables.
It turns out that these finite type cluster algebras were classified by S.~Fomin and A.~Zelevinsky~\cite{FominZelevinsky-ClusterAlgebrasII} using the Cartan-killing classification for crystallographic root systems.
Define the \defn{Cartan companion} of an exchange matrix $\B$ as the matrix~$\A{\B}$ given by~$a_{xy} = 2$ if~$x = y$ and~$a_{xy} = -|b_{xy}|$ otherwise.

@@

\para{Mutations in the $\b{g}$-vector fan}
%
We now discuss the linear dependences between $\b{g}$-vectors of adjacent seeds.
The following proposition was stated in~\cite[Lem.~19]{HohlwegPilaudStella}.

@@

Note that \cref{lem:linearDependencegvectorsCA} implies \cref{thm:generalizedAsso} since it ensures that the type cone~$\typeCone(\gvectorFan[\B_\circ])$ of the $\b{g}$-vector fan contains the cone of~${\b{h} \in \R^{\variables}}$ such that
\(
\b{h}_x + \b{h}_{x'} > \sum_{{y \in \cluster \cap \cluster', \; b_{xy} < 0}} -b_{xy}
\)
and
\(
% \quad\text{and}\quad
\b{h}_x + \b{h}_{x'} > \sum_{{y \in \cluster \cap \cluster', \; b_{xy} > 0}} b_{xy} \, \b{h}_y
\)
for any seeds~$(\B, \coefficients, \cluster)$ and~$(\B', \coefficients', \cluster')$ with~${\cluster \ssm \{x\} = \cluster' \ssm \{x'\}}$.

Unfortunately, \cref{lem:linearDependencegvectorsCA} is less precise than \cref{prop:exchangeablePairsAsso}. Indeed, which of the two possible linear dependences is satisfied by the $\b{g}$-vectors of~$\cluster \cup \cluster'$ depends on the initial exchange matrix~$\B_\circ$.
We will prove however that it is independent of the seeds containing~$x$ and~$x'$.

\begin{proposition}
\label{prop:uniqueExchangePropertyCA}
For any finite type exchange matrix~$\B_\circ$, the $\b{g}$-vector fan has the unique exchange relation property.
\end{proposition}

\subsubsection{Type cone of $\b{g}$-vector fans}

The linear dependences of \cref{lem:linearDependencegvectorsCA} provide a redundant description of the type cone of the $\b{g}$-vector fan~$\gvectorFan[\B_\circ]$.
We denote by~$\b{n}(\B_\circ, x, x')$ the normal vector of the inequality of the type cone corresponding to two exchangeable cluster variable~$x$ to~$x'$ (it is well defined by \cref{prop:uniqueExchangePropertyCA}.
In other words, depending on which of the two linear dependences of \cref{lem:linearDependencegvectorsCA} holds, we have
\[
\b{n}(\B_\circ, x, x') \eqdef \b{f}_x + \b{f}_{x'} - \sum_{\substack{y \in \cluster \cap \cluster' \\ b_{xy} < 0}} -b_{xy} \, \b{f}_y
\qquad\text{or}\qquad
\b{n}(\B_\circ, x, x') \eqdef \b{f}_x + \b{f}_{x'} - \sum_{\substack{y \in \cluster \cap \cluster' \\ b_{xy} > 0}} b_{xy} \, \b{f}_y,
\]
where~$(\b{f}_x)_{x \in \variables}$ denotes the canonical basis of~$\R^{\variables}$.
We obtain the following statement.

\begin{corollary}
\label{coro:typeConeCA}
For any finite type exchange matrix~$\B_\circ$, the type cone of the $\b{g}$-vector fan~$\gvectorFan[\B_\circ]$ is given by
\[
\typeCone(\gvectorFan[\B_\circ]) = \set{\b{h} \in \R^{\variables}}{\dotprod{\b{n}(\B_\circ, x, x')}{\b{h}} > 0 \text{ for all exchangeable cluster variables } x, x'}.
\]
\end{corollary}

In order to describe the facets of this type cone, we need the following special mutations.

\begin{definition}
Consider a seed~$(\B, \coefficients, \cluster)$ and a cluster variable~$x \in \cluster$. The mutation of~$\seed$ in direction~$x$ is a \defn{mesh mutation} if the entries~$b_{xy}$ for~$y \in X$ are all non-negative, or all non-positive.
\end{definition}

For instance, in the initial seed~$(\B_\circ, \coefficients_\circ, \cluster_\circ)$, the mutation in the direction of any variable~${x_\circ \in \cluster_\circ}$ is a mesh.
% and the corresponding $\b{g}$-vector linear dependence is~$\gvectorFull{\B_\circ}{x_\circ} + \gvectorFull{\B_\circ}{x_\circ'} = 0$.
%%, so that~$\b{n}(\B_\circ, x_\circ, x_\circ') = \b{f}_{x_\circ} + \b{f}_{x_\circ'}$. 
We call these meshes \defn{initial}.
The following statement describes the other meshes.

\begin{lemma}
\label{lem:bijectionMeshMutations}
There is precisely one non-initial mesh directed by each positive $\b{c}$-vector of the cluster algebra. Therefore, the number of non-initial meshes is~$N-n$.
\end{lemma}

%The linear dependence among the $\b{g}$-vectors of an initial mesh is just~$\gvectorFull{\B_\circ}{x_\circ} + \gvectorFull{\B_\circ}{x_\circ'} = 0$.
The following statement describes the linear dependence in any other mesh mutation.

\begin{lemma}
Consider two adjacent seeds~${(\B, \coefficients, \cluster)}$ and~${(\B', \coefficients', \cluster')}$ with~$\cluster \ssm \{x\} = \cluster' \ssm \{x'\}$ such that the entries~$b_{xy}$ for~$y \in X$ are all non-positive or all non-negative.
If either~$x$ or~$x'$ is an initial cluster variable, then
\[
\gvectorFull{\B_\circ}{x} + \gvectorFull{\B_\circ}{x'} = 0.
\]
Otherwise, the $\b{g}$-vectors of~$\cluster \cup \cluster'$ with respect to~$\B_\circ$ satisfy the linear dependence
\[
\gvectorFull{\B_\circ}{x} + \gvectorFull{\B_\circ}{x'} = \sum_{y \in \cluster \cap \cluster'} |b_{xy}| \, \gvectorFull{\B_\circ}{y}.
\]
%\[
%\b{n}(\B_\circ, x, x') = \b{f}_x + \b{f}_{x'} - \sum_{y \in \cluster \cap \cluster'} |b_{xy}| \, \b{f}_y
%\]
\end{lemma}

Our main result states that the linear dependences between the $\b{g}$-vectors of any mutation can be decomposed into positive combinations of linear dependences arising from meshes.

\begin{proposition}
\label{prop:extremalExchangeablePairsCA}
\end{proposition}

Our next statement follws from~\cref{prop:extremalExchangeablePairsCA,lem:bijectionMeshMutations}.

\begin{corollary}
\label{coro:simplicialTypeConeCA}
The type cone~$\typeCone(\gvectorFan[n])$ is simplicial.
\end{corollary}

Combining \cref{coro:simplicialTypeCone,coro:simplicialTypeConeCA,prop:extremalExchangeablePairsCA}, we derive the following description of all polytopal realizations of the $\b{g}$-vector fan~$\gvectorFan[\B_\circ]$. This result was stated in~\cite{BazierMatteDouvilleMousavandThomasYildirim} in the special situation of acyclic seeds in simply-laced types.

\begin{theorem}
For any finite type exchange matrix~$\B_\circ$, and for any~$\b{\ell} \in \R_{>0}^{N-n}$, the polytope
\[
R_{\b{\ell}}(\B_\circ) \eqdef \set{\b{z} \in \R^{\variables}}{...}
\]
is a generalized associahedron, whose normal fan is the $\b{g}$-vector fan~$\gvectorFan[\B_\circ]$.
\end{theorem}
